### 2023-07-11 intro.tex
Updates from Overleaf

Consider, the example ill-typed program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
\vspace{-3px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot.png}
\end{center}
\vspace{-3px}
A type checker with no support for error localization or recovery---%
as students in an undergraduate course might write---%
would simply report that this program is ill-typed. 
A more practical approach, and one common even in production type checkers, 
is to localize the first error that causes local type inference to fail and emit an explanatory error message before terminating.
In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then terminate.

An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018merlin}, would 
 be tasked with continuing past this first error.
 The general difficulty here is that there is now missing semantic information, namely the type of \li{f}, that 
 the bidirectional type system as specified 
 would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against.
 Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
it is unbound, so a second error can be simultaneously localized despite the missing information about \li{f}. Elm's compiler stops here.

To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Int} and \li{Bool}. 
There are several ways to localize this error. 
One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g. 
by observing recent editing history \cite{steady-typing}, 
or training a machine learning model \cite{SeidelBlame}. 
The other approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ, as indicated above. When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{potter2020hazel}---shows:

@@

This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
If localization had assumed that the \li{then} branch correct, as for example in Elm and \li{merlin}, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
If the \li{else} branch was chosen, then \li{x : Int} and an error would not be reported. 
If the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error here, 
so we choose not to report an error in Hazel.
No matter how localization handles the left operand, we would like to be able to recover and 
localize the type inconsistency on the right operand (the \li{+} operator is integer addition in Hazel).

This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to whether they make \emph{ad hoc} guesses about intent;  
(2) when combined with error recovery, one localization decision can influence downstream localization decisions; and 
(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of the topic.

>>>>

Consider, the example ill-typed program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
\vspace{-3px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot-v3.png}
\end{center}
\vspace{-3px}
A type checker with no support for error localization or recovery---%
as students in an undergraduate course might write---%
would simply report that this program is ill-typed. 
A more practical approach, and one common even in production type checkers, 
is to localize the first error that causes local type inference to fail and emit an explanatory error message before terminating.
In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then terminate.

An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018merlin}, would 
 be tasked with continuing past this first error.
 The general difficulty here is that there is now missing semantic information, namely the type of \li{f}, that 
 the bidirectional type system as specified 
 would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against. 
 Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
it is unbound, so a second error can be simultaneously localized despite the missing information.

To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Bool} and \li{Int}. 
There are several ways to localize this error. 
One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g. 
by observing recent editing history \cite{steady-typing}\todo{fix cite}, 
or training a machine learning model \cite{SeidelBlame}. 
The other approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ, as indicated above. When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020hazel}---shows:

@@

This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
If localization had assumed that the \li{then} branch correct, as for example in \li{merlin}, then 
\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
If the \li{else} branch was chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
 In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
If the inconsistency is localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error at either use of \li{x}, 
so we choose not to report an error in Hazel (and in fact treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing).
In any of these cases, we would like to be able to recover and 
localize the type inconsistency on the right addend because the \li{+} operator is integer addition in Hazel.
For the same reason, there is no definitive inconsistency with the branch types in the second conditional, so no 
error is marked there.

This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to whether they make \emph{ad hoc} guesses about intent;  
(2) when combined with error recovery, one localization decision can influence downstream localization decisions; and 
(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of these topics.

### 2023-07-11 let.tex
Changes from Overleaf
Updates from Overleaf

To ease the use of products, many languages feature destructuring bindings. In a typed setting, we may want to granularly add type annotations in patterns as well. As it turns out this combination of features is surprisingly tricky to get right! Let us add let
expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%

@@

The marking rule is directly analogous and utilizes the marking of patterns, which we omit---they
may be derived in the same way as those for expressions. In the analytic cases, we introduce error
marks that parallel to $\ECInconType{\ECMV}$ and $\ECAnaNonMatchedProd{\ECMV}$.
\[%
  \judgment{
    \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynFixedInto{\ctx'}{\EMV_{2}}{\ECMV_{2}}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
  }{ISLetPat}
\]%

>>>>

To ease the use of products and other datatypes, many languages feature destructuring bindings. In a typed setting, we may want to granularly add type annotations in patterns as well. As it turns out this combination of features is surprisingly tricky to get right! Let us add let
expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%

@@

The marking rule is directly analogous and utilizes the marking of patterns, which we omit---they
may be derived in the same way as those for expressions and are governed by similar metatheorems. 
In the analytic cases, we introduce error marks that parallel to $\ECInconType{\ECMV}$ and 
$\ECAnaNonMatchedProd{\ECMV}$.
\[%
  \judgment{
    \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynFixedInto{\ctx'}{\EMV_{2}}{\ECMV_{2}}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
  }{ISLetPat}
\]%
